Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2022-05-17 analysis/paper/intro.Rmd
further development of manuscript
push manuscript files to github

# Introduction

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus, mineralised dental plaque, has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerDirectEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculusCooking2014]. 
It has also proven to be able to retain and preserve dietary information for millenea,
providing detailed dietary information on past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

<!-- more detailed background, comprehensible to scientists in related disciplines -->

Until recently, few studies directly investigated the presence of plant microremains
from the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentificationOpal1975], 
and later by Middleton and Rovner [-@middletonExtractionOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. The ability to detect coca
consumption by looking at human dental calculus was explored by Klepinger and Kuhn
[-@klepingerCalculusCoca1977], and starches and phytoliths were extracted from 
human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].

In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has been, 
to a larger extent, acknowledged. The applications of dental calculus span a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncientCalcified2013; @warinnerPathogensHostImmunity2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsightsConsumption2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculusCooking2014; @hendyProteomicCalculus2018], 
and nicotine-use [@eerkensDentalCalculusSource2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

<!-- starches from dental calculus -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways and the methods we use to 
extract and analyse dental calculus and make inferences on past diets. 
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. Other studies have been conducted on modern populations 
[@leonardDentalCalculus2015] and primates [@powerChimpCalculus2015], both with
well-documented dietary intake, to reconcile our interpretations on the recovery 
of microremains (both phytoliths and starches) from dental calculus and how well
they represent actual dietary intake. These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a
multifactorial aetiology and variable formation rates both within and between
individuals [@jepsenCalculusRemoval2011].
Additionally, the effects of the many different methods
of starch processing, as well as post-depositional processes, are still being 
explored [@graneroStarchTaphonomy2020].

<!-- clearly stating the general problem being addressed by this particular study -->
<!-- Dental calculus research has its limitations, all of which I will solve with this
single method. -->
In this exploratory study, we use an oral biofilm model to investigate the 
retention of starch granules within dental calculus, in a controlled laboratory 
setting. Our main questions concern the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is 
incorporated into the calculus, and how much is recovered? 
Is there differential loss of information from specific dietary markers that affects 
the obtained dietary information, and how does this affect the representation of diet from
extracted microremains? We find that a limited proportion of the starch input is actually 
retained in the calculus, as well as a shift in the size ratios of individual starch granules
that are incoporated into the calculus. We also show that the number of starch
granules that are incorporated increases as the size of the calculus deposit 
increases.

>>>>

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculus2014]. 
This dietary information can be preserved within the mineralised dental plaque 
over many millennia, providing a unique window into the food-related behaviours of 
past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentification1975], 
and later by Middleton and Rovner [-@middletonOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncient2013; @warinnerPathogensHost2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsights2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculus2014; @hendyProteomicCalculus2018], 
and nicotine use [@eerkensDentalCalculus2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodPathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015], 
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

In this exploratory study, we use an oral biofilm model to investigate the 
retention of starch granules within dental calculus in a controlled laboratory 
setting, allowing us full control over dietary input. Our main questions concern
the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is 
incorporated into the calculus, and how much is recovered? 
Is there differential loss of information from specific dietary markers that affects 
the obtained dietary information, and how does this affect the representation of 
diet from extracted microremains?  
We find that, despite the absence of $\alpha$-amylase in
the model, a limited proportion of the starch input is actually 
retained in the calculus. We also observed a shift in the size ratios of individual 
starch granules that are incorporated into the calculus, and that the number of 
incorporated starch granules increases as the size of the calculus deposit 
increases.

### 2022-05-17 analysis/paper/matmet.Rmd
further development of manuscript
push manuscript files to github

# Materials and Methods

```{r protocol-fig, fig.cap="Overview of experiment protocol"}
knitr::include_graphics(here("analysis/figures", "ExpL_protocol.png"))
```

<!-- remove sampling from protocol picture -->

@@

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues [-@sissonsMultistationDentalPlaque1991]. 
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationDentalPlaque1991], 6.4 mmol/l NaHCO~3~ 
[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
and 1 mmol/l arginine [@sissonsMultistationDentalPlaque1991].

<!-- add to protocols.io -->
Fresh whole saliva (WS) for inoculation was provided by a 31-year-old male donor 
with no history of caries, who abstained from oral hygiene for 24 hours, and no 
food was consumed two hours prior to donation. No antibiotics were taken up to six 
months prior to donation. 
The saliva was filtered through a sterilised (with bleach) nylon cloth to remove
particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
20% glycerine for four hours at 36&deg;C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
substrata were transferred to a new plate containing 1 ml/well AS and incubated 
at 36&deg;C, 30 rpm. The inoculation process was repeated on days 3 and 5. 
AS was partially refreshed once per day and fully refreshed every three days, 
throughout the experiment, by transferring the substrata to a new plate containing 
AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

Starch treatments were initiated on day 9 to avoid starch granule counts being 
affected by $\alpha$-amylase hydrolysis. An $\alpha$-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the system (see 
Bartholdy et al. ???? <!-- ## [bartholdy] ## --> for more details). Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a 
new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch solutions were created in a 5%
(w/v) sucrose solution. Before transferring biofilm samples to the starch treatments, 
the starch plates were agitated to keep the starches in suspension in the solutions, 
and during treatments, the rpm was increased to 60.

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991], 
and (0.04 g/l MgCl). The substrata were submerged in 1 ml/well CPMU five times daily, 
every two hours, for six minutes. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This cycle was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of protocol). More detailed protocols
are available on protocols.io<!--insert DOI-->.

All laboratory work was conducted in sterile conditions under a laminar flow hood 
to prevent starch and bacterial contamination. Control samples were included to 
detect starch contamination.

## Starch counts

Starch granule counts were divided into three size categories: small (<10 $\mu$m),
medium (10 -- 20 $\mu$m), and large (>20 $\mu$m). Counts for small wheat and
potato starches were combined for the mixed treatment samples, as it is not 
possible to distinguish the small starch granules between the two species.

### Treatment solutions

A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). Counting starches on a full slide was not feasible, so three slide transects were 
counted (at 1/3, 1/2, and 2/3 of the slide), and the sample counts were extrapolated 
(see Supplementary material for more details).

### Extraction from samples

Extraction of starches from the calculus samples was performed by dissolving the
calculus in ethylenediaminetetraacetic acid (EDTA), as it has been shown to
have little to no effect on starch granules (especially native granules)
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When dispersing the liquid (EDTA) sample on the slide, the sample was homogenised using the pipette. 
It was not possible to count all wheat starches
on a slide, so wheat and mix counts were extrapolated as described for the treatment
solutions.

@@

Statistical analysis was conducted in R versions 4.0.5 ("Shake and Throw") 
and 4.0.3 ("Bunny-Wunnies Freak Out") [@R-base].
<!-- R packages used -->
A one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
as the grouping variable (GV) was conducted to explore the effect of the different 
starch treatments on biofilm growth.

To calculate size proportions, mean counts for each treatment were taken across 
both experimental plates for each treatment, resulting in a mean count for each
granule size category within each treatment.

Pearson's r was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.

Scripts and data will be made available on osf.io upon publication.
Until then, they can be found on the first author's GitHub repository
(https://github.com/bbartholdy/byocStarch).

<!-- script and data availability: provide link (or doi?) to osf.io repository -->

>>>>

@@

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
were the substrata on which the calculus grew. Using this system allowed for easy
transfer of the growing biofilm between saliva, feeding solutions, 
and mineral solutions.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues [-@sissonsMultistationPlaque1991]. 
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationPlaque1991], 6.4 mmol/l NaHCO~3~ 
[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
and 1 mmol/l arginine [@sissonsMultistationPlaque1991].

Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
with no history of caries, who abstained from oral hygiene for 24 hours. No 
food was consumed two hours prior to donation and no antibiotics were taken up 
to six months prior to donation. 
The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
nylon cloth to remove particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
substrata were transferred to a new plate containing 1 ml/well AS and incubated 
in a shaking incubator (Infors HT Ecotron) at 36&deg;C, 30 rpm. 
The inoculation process was repeated on days 3 and 5. 
AS was partially refreshed once per day and fully refreshed every three days, 
throughout the experiment, by transferring the substrata to a new plate containing 
stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

Starch treatments were initiated on day 9 to avoid starch granule counts being 
affected by $\alpha$-amylase hydrolysis from saliva inoculation days. 
An $\alpha$-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the model before
starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a 
new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch treatments were created in dH~2~O
with 5% (w/v) sucrose. Before transferring biofilm samples to the starch treatment
plate, the plates were agitated to keep the starches in suspension in the
solutions. During treatments, the rpm was increased to 60 to facilitate contact
between starch granules and biofilms.

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).

All laboratory work was conducted in sterile conditions under a laminar flow hood 
to prevent starch and bacterial contamination. Control samples that only received
sucrose as a treatment were included to detect starch contamination from the
environment or cross-contamination from other wells in the same plate.

## Amylase activity detection

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing stock AS (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a modified version of an Enzymatic Assay of $\alpha$-Amylase
(<https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase>) [@bernfeldAmylase1955], which measures the amount of
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 $\mu$mole of maltose in 6 minutes.
The detailed protocol can be found here: <https://www.protocols.io/view/amylase-activity-bw8jphun>.

## Treatment solutions

A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three 
slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
were extrapolated to the total number of granules exposed to the samples over 16
days of treatments (see Supplementary Material for more details). For potato
treatment samples, the whole slide was counted.

## Extraction method

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material for more detail).

Both wheat and potato granules were divided into three size categories: 
small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).

@@

Statistical analysis was conducted in `r version$version.string`  [@Rbase] and
the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
here [@R-here], and patchwork [@R-patchwork].
<!-- R packages used -->

To see if biofilm growth was differently affected by starch treatments, a 
one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
as the grouping variable (GV) was conducted.
To analyse granule counts and calculate size proportions, mean counts for each 
treatment were taken across
both experimental plates, resulting in a mean count for each
granule size category within each treatment.

Pearson's *r* was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.
This was applied to total biofilm weight and starch count per mg 
calculus (also z-score standardised) to account for differences in starch 
concentration in the calculus [as per 
@wesolowskiEvaluatingMicrofossilContent2010].
